Streaming $2$-Center with Outliers in High Dimension

### 2015-04-15 paper.tex
revised introduction

\section{Introduction}
Metric $k$-center problem -- covering a set of point with $k$ congruent balls of minimum radius -- is a fundamental problem in computer science and has various applications in data mining and image processing.
One of $k$-center problem variants is $k$-center with $z$-outlier which finds $k$ congruent balls with minimum radius covering all but $z$ points of a point set. In real world applications, due to commonness of noisy data, it is important to consider outlier points. Because even a small number of outliers ($\cO(k)$) can increase the radius of $k$-center problem unboundedly.
In this paper we focus on streaming data model. In the streaming data model, input arrives as a sequence and there is not enough space to keep the whole data set. Algorithms in this model have limitations in space (typically sub-linear in input size) and processing time per item. This model is specially suitable for massive data sets, due the its sub-linear space complexity \cite{aggarwal2007data}.

In this paper our concern is about $2$-center problem in high dimensional Euclidean space.
The formal definition of problem is as follows. Given sequence $\mathcal{S}$ of points in $d$ dimensional (for arbitrarily large $d$) Euclidean space, find two congruent balls of minimum radius, covering all but at most $z$ points.

@@

>>>>

\section{Introduction}
Metric $k$-center problem -- covering a set of point with $k$ congruent balls of minimum radius -- is a fundamental problem in computer science and has various applications in data mining and image processing.
One of $k$\nobreakdash-center problem variants is $k$-center with $z$-outlier which finds $k$ congruent balls with minimum radius covering all but $z$ points of a point set. In real world applications, due to commonness of noisy data, it is important to consider outlier points. Because even a small number of outliers ($\cO(k)$) can increase the radius of $k$\nobreakdash-center problem unboundedly.
In this paper we focus on streaming data model. In the streaming data model, input arrives as a sequence and there is not enough space to keep the whole data set. Algorithms in this model have limitations in space (typically sub-linear in input size) and processing time per item. This model is specially suitable for massive data sets, due the its sub-linear space complexity \cite{aggarwal2007data}.

In this paper our concern is about $2$-center problem in high dimensional Euclidean space.
The formal definition of problem is as follows. Given sequence $\mathcal{S}$ of points in $d$ dimensional Euclidean space (for arbitrarily large $d$), find two congruent balls of minimum radius, covering all but at most $z$ points.

$k$-center problem without outliers (i.e., $z=0$) in general (Euclidean) space can not be approximated with a smaller factor than $2$ ($1.822$), unless $\mathbf{P}=\mathbf{NP}$ \cite{bern1996approximation}.
Gonzalez showed a $2$-approximation algorithm for this problem \cite{gonzalez1985clustering}.
For small $k$ and $d$, there are more efficient algorithms.
For Euclidean $1$-center in fixed dimensions, Chazelle~\etal{} gave a linear algorithm \cite{chazelle1996linear}.
For $2$\nobreakdash-center in the plane the best known algorithm, given by Chan \cite{chan1999more}, runs in $\cO(n \log^2 n \log^2 \log n)$ and linear space.

In streaming data model, Guha \cite{guha2009tight}, in parallel with McCutchen~\etal{} \cite{mccutchen2008streaming} gave $(2+\eps)$-approximation for $k$-center without outliers.
For small $k$, better algorithms are known. For fixed $d$, there exists an $\eps$\nobreakdash-coreset for $1$-center with $\cO(\eps^{-\frac{d-1}{2}})$ space.
For higher dimensions, Agarwal \cite{agarwal2010streaming} gave a $(\frac{1+\sqrt{3}}{2})$-approximation for $1$\nobreakdash-center problem, and Chan \cite{chan2004faster} reduced its approximation factor to $1.22$, by an improved analysis.
For $k=2$, Ahn~\etal{}~\cite{ahn2014computing} gave a $(2+\eps)$-approximation algorithm using $\cO(\frac{d}{\eps})$ space, which was improved for Euclidean spaces by Kim~\etal{} \cite{kim2014improved} to a $(1.8+\eps)$-approximation algorithm, using the same space.

For the problem of $k$-center with outliers, Charikar~\etal{} \cite{charikar2001algorithms} gives the first offline algorithm with approximation factor $3$ ($4$), where the candidate center points of the centers can be enumerated (otherwise).
For small $k$, there are better results known. $1$-center with $z$ outliers can be solved in $\cO(n \log n + z^{3} n^{\eps})$, for any $\eps > 0$, using Matou{\v{s}}ek's framework \cite{matouvsek1995geometric}.
Agarwal \cite{agarwal2008efficient} gives a randomized $\cO(n z^{7} \log^3 z)$-time algorithm for $2$-center problem in $\IR^2$.

For the problem of $k$-center with $z$ outliers in streaming data model, McCutchen~\etal{} \cite{mccutchen2008streaming} (using idea in Charikar~\etal{} \cite{charikar2001algorithms} as subroutine) give a $(3+\eps)$-approximation ($(4+\eps)$-approximation, depending on the used subroutine) algorithm using $\cO(\frac{zk}{\eps})$ space.
In special case of $k=1$ in fixed dimensions, there is a streaming $\cO(\frac{z}{\eps^{\cO(d)}})$-space algorithm \cite{agarwal2007space, zarrabi2011almost}.
For Euclidean spaces, Zarrabi-Zadeh \cite{zarrabi2009streaming} gave $(1.22 \sqrt{2})$\nobreakdash-approximation for $1$-center with $z$ outliers running in $\cO(?)$ space.

In this paper we first give a simple $2$-approximation algorithm for $1$-center problem with space complexity $\cO(z^2 + zd)$. We then show a $(1.8 + \eps)$-approximation streaming algorithm for $2$-center with $z$ outliers in Euclidean space, with space complexity polynomial in $z$ and $d$ and sub-linear in $n$.

@@

%---------------------------- Conclusions ---------------------

\section{Conclusions}
\label{sec:conc}

In this paper, we presented a $(1.8 + \eps)$-approximation streaming algorithm for $2$-center problem with outliers in Euclidean space, which improves previous $(3+\eps)$-approximation algorithm by Charikar~\etal~\cite{mccutchen2008streaming}. It remains open whether there can be any improvements in approximation factor. It is an interesting problem to find an algorithm with better approximation factor or space complexity. It is also interesting to see if the ideas in this paper can be applied to $k$-center problem with outliers in streaming data model, even for small $k$.
